Multitask matrix completion for learning protein interactions across diseases

### 2016-09-01 paper.tex
new pdf

%\usepackage[top=1.3in, bottom=1.3in, left=1in, right=1in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}
\usepackage{subfigure} 
\usepackage{natbib}
%\usepackage[footnotesize]{caption}
\usepackage{multirow}
\usepackage{url}

@@

\subsection{Background: Host-pathogen protein interactions}
%An important part of our sustenance depends on how we counter infectious diseases -- those that are caused by external agents like pathogens. To understand how these microbial organisms reside and disrupt the host processes we need to comprehend the cross-organism protein interactions between the host species such as mammals, and the causative pathogen species such as viruses and bacteria. %These interactions  pathogen proteins physically bind with human proteins to manipulate important biological processes in the host cell, evade the host immune response and multiply within the host. 
The experimental discovery of host-pathogen protein interactions involves biochemical and biophysical methods such as co-immunoprecipitation (co-IP), yeast two-hybrid (Y2H) assays, co-crystalization. The host-pathogen protein interactions from several small-scale and high throughput experiments are aggregated by databases such as virus-MINT\cite{virusmint}, HPIDB \cite{hpidb}, PHISTO \cite{phisto} etc by literature curation. These databases are valuable sources of information to bioinformaticians for developing models. \\

\noindent\textbf{Prediction of host-pathogen PPIs}: The most reliable experimental methods for studying protein-protein interactions (PPI) are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- for example, the bacterium \textit{Bacillus anthracis} which causes anthrax has about 2321 proteins which when coupled with the 100,000 or so human proteins gives $\approx$232 million protein pairs to test, experimentally. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. These techniques use the known interactions data from previous experiments and predict the most plausible new interactions. In particular, supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins such as: protein sequences from UniProt \cite{UniProt}, protein structure from PDB, gene ontology from GO database \cite{GO}.

@@

\subsection{Comparison with other machine learning approaches}
\label{mlmethods}
Our baselines include recent low-rank and sparse models, conventional multitask methods and prior work on HP PPI prediction. For a uniform comparison we used least squared loss in all the methods. The MALSAR \cite{malsar} package was used to implement some of the models. For the baselines wherever appropriate, we concatenated the features of the two node types into a single feature vector. Let $W \in \mathbb{R}^{T \times d_t}$ be the matrix with the task-specific weight vectors $\mathbf{w}_t$.

@@

\noindent\textit{Performance on the homologous PPI partition}:
Except the \textit{Ebola} task, BSL-MTL has the highest AUC. The Homolog baseline best captures the homologous PPI in \textit{Ebola} (with an AUC twice the next best). However, since there are far fewer homologous PPI in Ebola than in the other tasks (row-2 of Table \ref{datasets}), this contributes less to the overall performance of the Homolog baseline. \textit{Hepatitis-C} has the highest fraction of homologous PPI and BSL-MTL does significantly better on these than the other baselines which also leads to a higher overall AUC. The same can be said of the \textit{Influenza} task.

@@

\begin{table}[bht]
\label{goterms}
\begin{scriptsize}
\begin{center}
%\def\arraystretch{1.1}
\begin{tabular}{|c|c|}
\multicolumn{2}{c}{Gene Ontology terms for clusters from Figure \ref{fig:pca2}} \\ \hline
cluster 1 & mRNA (guanine-N7-)-methyltransferase activity; RNA-directed \\
(ebola only)& RNA polymerase activity; ATP binding \\ \hline
cluster 2 & host cell endoplasmic reticulum membrane; \\ 
(hepatitis \& & host cell lipid particle; induction by virus \\
influenza & of host autophagy; host cell mitochondrial membrane; \\
proteins) & suppression by virus of host STAT1 activity \\ \hline
cluster 3 (all) & fusion of virus membrane with host plasma membrane \\ \hline
\end{tabular}
\caption{GO terms (for process, function, component) enriched in the highlighted clusters from Figure \ref{fig:pca2}. FuncAssociate GO enrichment tool was used and terms with p-value < 0.001 are shown. The first column also indicates the composition of each cluster (i.e which task's proteins appear in the cluster).}
\end{center}
\end{scriptsize}
\end{table}

>>>>

%\usepackage[top=1.3in, bottom=1.3in, left=1in, right=1in]{geometry}
\usepackage{floatrow}
\usepackage{graphicx} % more modern
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{subfigure} 
\usepackage{natbib}
%\usepackage[footnotesize]{caption}
\usepackage{multirow}
\usepackage{url}

@@

\subsection{Background: Host-pathogen protein interactions}
%An important part of our sustenance depends on how we counter infectious diseases -- those that are caused by external agents like pathogens. To understand how these microbial organisms reside and disrupt the host processes we need to comprehend the cross-organism protein interactions between the host species such as mammals, and the causative pathogen species such as viruses and bacteria. %These interactions  pathogen proteins physically bind with human proteins to manipulate important biological processes in the host cell, evade the host immune response and multiply within the host. 
The experimental discovery of host-pathogen protein interactions involves biochemical and biophysical methods such as co-immunoprecipitation (co-IP), yeast two-hybrid (Y2H) assays, co-crystalization. The host-pathogen protein interactions from several small-scale and high throughput experiments are aggregated by databases such as MINT \citep{virusmint}, HPIDB \citep{hpidb}, PHISTO \citep{phisto} etc by literature curation. These databases are valuable sources of information to bioinformaticians for developing models. \\

\noindent\textbf{Prediction of host-pathogen PPIs}: The most reliable experimental methods for studying protein-protein interactions (PPI) are often very time-consuming and expensive, making it hard to investigate the prohibitively large set of possible host-pathogen interactions -- for example, the bacterium \textit{Bacillus anthracis} which causes anthrax has about 2321 proteins which when coupled with the 100,000 or so human proteins gives $\approx$232 million protein pairs to test, experimentally. Computational techniques complement laboratory-based methods by predicting highly probable PPIs. These techniques use the known interactions data from previous experiments and predict the most plausible new interactions. In particular, supervised machine learning based methods use the few known interactions as training data and formulate the interaction prediction problem in a classification setting, with target classes: ``interacting'' or ``non-interacting''. Features are derived using various attributes of the two proteins such as: protein sequences from UniProt \citep{UniProt}, protein structure from PDB, gene ontology from GO database \citep{GO}.

@@

\subsection{Comparison with other machine learning approaches}
\label{mlmethods}
Our baselines include recent low-rank and sparse models, conventional multitask methods and prior work on HP PPI prediction. For a uniform comparison we used least squared loss in all the methods. The MALSAR \citep{malsar} package was used to implement some of the models. For the baselines wherever appropriate, we concatenated the features of the two node types into a single feature vector. Let $W \in \mathbb{R}^{T \times d_t}$ be the matrix with the task-specific weight vectors $\mathbf{w}_t$.

@@

\noindent\textit{Performance on the homologous PPI partition} (Table \ref{homresultsTable}):
Except the \textit{Ebola} task, BSL-MTL has the highest AUC. The Homolog baseline best captures the homologous PPI in \textit{Ebola} (with an AUC twice the next best). However, since there are far fewer homologous PPI in Ebola than in the other tasks (row-2 of Table \ref{datasets}), this contributes less to the overall performance of the Homolog baseline. \textit{Hepatitis-C} has the highest fraction of homologous PPI and BSL-MTL does significantly better on these than the other baselines which also leads to a higher overall AUC. The same can be said of the \textit{Influenza} task.

@@

\begin{table}[bht]
\begin{scriptsize}
\begin{center}
%\def\arraystretch{1.1}
\begin{tabular}{|c|c|}
\multicolumn{2}{c}{Gene Ontology terms for clusters from Figure \ref{fig:pca2}} \\ \hline
cluster 1 & mRNA (guanine-N7-)-methyltransferase activity; RNA-directed \\
(ebola only)& RNA polymerase activity; ATP binding \\ \hline
cluster 2 & host cell endoplasmic reticulum membrane; \\ 
(hepatitis \& & host cell lipid particle; induction by virus \\
influenza & of host autophagy; host cell mitochondrial membrane; \\
proteins) & suppression by virus of host STAT1 activity \\ \hline
cluster 3 (all) & fusion of virus membrane with host plasma membrane \\ \hline
\end{tabular}
\caption{GO terms (for process, function, component) enriched in the highlighted clusters from Figure \ref{fig:pca2}. FuncAssociate GO enrichment tool was used and terms with p-value < 0.001 are shown. The first column also indicates the composition of each cluster (i.e which task's proteins appear in the cluster).}
\label{goterms}
\end{center}
\end{scriptsize}
\end{table}
